 

### 2022-05-10 thesis/introduction.qmd
Move thesis from R Markdown to Quarto

---
title: ''
output: pdf_document
bibliography: ../references.bib
---

@@

Measured breath-by-breath data is noisy (see grey points in Figure 1). Both the biological variability of breathing patterns and the measurement error --- as well as irregular breaths such as coughs and swallowing --- lead to a highly fluctuating raw oxygen uptake. For interpretation the raw data requires some form of processing.

@@

![Raw breath by breath data processed by different strategies during the final minutes of a ramp test to exhaustion. Grey points display oxygen uptake from the single breaths. The moving averages were calculated over 30 breaths and seconds, respectively. The binned average was calculated over 30-second intervals with the mean values aligned to the center of each interval. For digital filtering a forward-backward (zero phase) low-pass Butterworth filter with the parameters 0.04 Hz (cut-off frequency) and 3 (order) for each filter was applied. For more details on the used data processing strategies, see the methods section.](../plots/example.png)

>>>>

---
title: ''
format: pdf
bibliography: ../references.bib
---

@@

Measured breath-by-breath data is noisy (see grey points in @fig-example). Both the biological variability of breathing patterns and the measurement error --- as well as irregular breaths such as coughs and swallowing --- lead to a highly fluctuating raw oxygen uptake. For interpretation the raw data requires some form of processing.

@@

![Raw breath by breath data processed by different strategies during the final minutes of a ramp test to exhaustion. Grey points display oxygen uptake from the single breaths. The moving averages were calculated over 30 breaths and seconds, respectively. The binned average was calculated over 30-second intervals with the mean values aligned to the center of each interval. For digital filtering a forward-backward (zero phase) low-pass Butterworth filter with the parameters 0.04 Hz (cut-off frequency) and 3 (order) for each filter was applied. For more details on the used data processing strategies, see the methods section.](../plots/example.png){#fig-example}

### 2022-05-10 thesis/methods.qmd
Move thesis from R Markdown to Quarto

I conducted all analyses using R Version 4.1.2 [@rcoreteam2022]. The thesis was entirely written in R Markdown Version 1.1 [@allaire2022]. The attached packages and default settings are documented in Appendix A.4.

@@

The article search was conducted on 16th March 2022 using [PubMed]() and [Web of Science](). The search included articles published from 2017 to 2022 referring to 'maximum oxygen uptake' or equivalent terms in title, abstract or keywords. Table 1 shows the exact search terms used.

```{r terms, echo = FALSE}
out <- data.frame(
  source = c("PubMed", "Web of Science"),
  term = c('`((((("maximum oxygen uptake") OR ("maximal oxygen uptake")) OR ("VO2max")) OR ("maximum oxygen consumption")) OR ("maximal oxygen consumption")) AND (("2017/01/01"[Date - Publication] : "3000"[Date - Publication]))`', '`(((((ALL=("maximum oxygen uptake")) OR ALL=("maximal oxygen uptake")) OR ALL=("VO2max")) OR ALL=("maximum oxygen consumption")) OR ALL=("maximal oxygen consumption")) AND PY=(2017-2022)`')
)

knitr::kable(out, 
  col.names = c("Source", "Search String"), 
  caption = "Search strings for the systematic scoping review."
)
```

The search results from both data bases were merged and checked for the presence of a Digital Object Identifier (DOI). Entries without DOI were excluded to allow for automated removal of duplicates by DOI matching in the next step. After removing the duplicates I conducted an automated title scanning to exclude results that were likely no original research articles. Table 2 displays the exclusion terms for the automated title screening.

```{r titlescan, echo = FALSE}
out <- data.frame(
  term = c("review, correction, meta-analysis, comment, retraction, editorial, erratum, reply")
)

knitr::kable(out, 
  col.names = c("Search terms for exclusion by title"), 
  caption = "Exclusion terms for the automated title screening after removal of duplicates. If the title matched at least one of the given terms, the article was excluded.")
```

@@

The abstracts from the articles included in the random sample were blinded for scanning. This meant removing any further information not relevant for the screening---such as authors or journal---leaving only the title, abstract and an ID of the article (see Appendix A.3 for an example). Two researchers independently scanned the abstracts to filter those that matched one of the exclusion criteria shown in Table 3. When the screeners disagreed in their assessments, they resolved the conflict by discussion.

```{r exclusion, echo = FALSE}
exclusion_criteria <- data.frame(
  id = c("A*", "B*", "C*", "D*", "E", "F"),
  criteria = c("not in English", "no primary research", "research not in humans", "VO2max only estimated", "no appropriate test protocol", "no information regarding the exclusion criteria"),
  details = c("Full text only available in non-English language", "research was no original investigation or only a reanalysis of data", "research was conducted in animals", "VO2max was only approximated by means of a predictive equation", "Protocol for VO2max testing did either not include exercise to voluntial exhaustion or was to long (>20 min) for a reliable estimate", "crucial information on VO2max testing that allowed the evaluation of the other exclusion criteria were missing")
)

knitr::kable(
  exclusion_criteria,
  col.names = c("", "Criterium", "Details"),
  caption = "Exclusion criteria for the screening process. During the abstract screening only the criteria marked with an asterix were evaluated; during full text screening all criteria were assessed."
)
```

After the abstract screening I retrieved the full texts for all articles remaining in the review. The full texts were again independently scanned by two researchers to include only those articles that measured V̇O~2max~ using an appropriate testing procedure in humans (see Table 3 for the detailed full-text exclusion criteria). Conflicts were resolved by discussion.

>>>>

I conducted all analyses using R Version 4.1.2 [@rcoreteam2022] within R Studio Version 2022.2.2.485 [@rstudioteam2022]. The thesis was entirely written in Quarto Version 0.9.380 [@allaire2022a] . The attached packages and default settings are documented in Appendix A.4.

@@

The article search was conducted on 16th March 2022 using [PubMed]() and [Web of Science](). The search included articles published from 2017 to 2022 referring to 'maximum oxygen uptake' or equivalent terms in title, abstract or keywords. @tbl-search shows the exact search terms used.

| Source         | Search String                                                                                                                                                                                                            |
|----------------|--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| PubMed         | `((((("maximum oxygen uptake") OR ("maximal oxygen uptake")) OR ("VO2max")) OR ("maximum oxygen consumption")) OR ("maximal oxygen consumption")) AND (("2017/01/01"[Date - Publication] : "3000"[Date - Publication]))` |
| Web of Science | `(((((ALL=("maximum oxygen uptake")) OR ALL=("maximal oxygen uptake")) OR ALL=("VO2max")) OR ALL=("maximum oxygen consumption")) OR ALL=("maximal oxygen consumption")) AND PY=(2017-2022)`                              |

: Search strings for the systematic scoping review. {#tbl-search}

The search results from both data bases were merged and checked for the presence of a Digital Object Identifier (DOI). Entries without DOI were excluded to allow for automated removal of duplicates by DOI matching in the next step. After removing the duplicates I conducted an automated title scanning to exclude results that were likely no original research articles. @tbl-titlescan displays the exclusion terms for the automated title screening.

| Search terms for exclusion by title                                               |
|-----------------------------------------------------------------------------------|
| review, correction, meta-analysis, comment, retraction, editorial, erratum, reply |

: Exclusion terms for the automated title screening after removal of duplicates. If the title matched at least one of the given terms, the article was excluded. {#tbl-titlescan}

@@

The abstracts from the articles included in the random sample were blinded for scanning. This meant removing any further information not relevant for the screening---such as authors or journal---leaving only the title, abstract and an ID of the article (see Appendix A.3 for an example). Two researchers independently scanned the abstracts to filter those that matched one of the exclusion criteria shown in @tbl-exclusion. When the screeners disagreed in their assessments, they resolved the conflict by discussion.

|     | Criterium                                       | Details                                                                                                                                 |
|-----|-------------------------------------------------|-----------------------------------------------------------------------------------------------------------------------------------------|
| A\* | not in English                                  | Full text only available in non-English language                                                                                        |
| B\* | no primary research                             | research was no original investigation or only a reanalysis of data                                                                     |
| C\* | research not in humans                          | research was conducted in animals                                                                                                       |
| D\* | V̇O~2max~ only estimated                         | V̇O~2max~ was only approximated by means of a predictive equation                                                                        |
| E   | no appropriate test protocol                    | Protocol for V̇O~2max~ testing did either not include exercise to voluntial exhaustion or was to long (\>20 min) for a reliable estimate |
| F   | no information regarding the exclusion criteria | crucial information on V̇O~2max~ testing that allowed the evaluation of the other exclusion criteria were missing                        |

: Exclusion criteria for the screening process. During the abstract screening only the criteria marked with an asterix were evaluated; during full text screening all criteria were assessed. {#tbl-exclusion}

After the abstract screening I retrieved the full texts for all articles remaining in the review. The full texts were again independently scanned by two researchers to include only those articles that measured V̇O~2max~ using an appropriate testing procedure in humans (see @tbl-exclusion for the detailed full-text exclusion criteria). Conflicts were resolved by discussion.

### 2022-05-10 thesis/results.qmd
Move thesis from R Markdown to Quarto

Initial search yielded 7529 results of which 4364 remained after automated filtering and removal of duplicates (see flow diagram in Figure 1). Out of the random sample (n = 500), 242 articles were included in the final analysis.

![Flow diagram for the systematic scoping review in accordance with the PRISMA 2020 Statement [@page2021]](../plots/flow_diagram.png)

Reporting practices of the methodology of gas exchange measures differed widely across the literature (see Table 4). More than half (51.8%) of the articles did not report any information regarding their data processing strategy. One in twenty articles (5.4%) provided a rationale for their used strategy. Only a single article [@maturana2021] reported information regarding all the investigated criteria.

```{r reporting-table, echo=FALSE}
load("../data/reporting.Rda")
t <- purrr::map_dfr(reporting, scales::percent, accuracy = 0.1)

@@

knitr::kable(
  t,
  col.names = c("Metabolic cart", "Preprocessing", "Software", "Processing Strategy", "Reference"),
  caption = "Percentage of studies that provided details on the different characteristics of oxygen uptake data processing. *only examined within the subgroup of studies using breath-by-breath measurements",
  align = "r"
)
```

Out of the authors that provided information and collected breath-by-breath measurements, most (79.5%) utilized binned averages to determine V̇O~2max~. Moving time averages, or breath-based averages were uncommon (see Figure 2). No study used methods of digital filtering to determine V̇O~2max~.

![Data strategies for processing breath-by-breath data in the reviewed literature (n = 88).](../plots/strategies.png)

@@

The calculation interval for time-based averages of mixing chamber and breath-by-breath devices ranged from 5 to 60 seconds (see Figure 3). 30 second length intervals were most common to define V̇O~2max~, while authors also often employed shorter (10-20 s) and longer (60 s) periods.

![Total durations of the calculation interval of V̇O~2max~ in the reviewed studies.](../plots/duration_count.png)

## Experimental Comparison

The average V̇O~2max~ as determined by a binned 30-second average was 62.2±6.3 ml·min^-1^·kg^-1^ (mean ± standard deviation). Applying different data processing strategies for V̇O~2max~ determination leads to different outcome values (see Figure 4).

![V̇O~2max~ varies by data processing strategy. Values are expressed relative to the V̇O~2max~ from a 30-second binned average --- the most common strategy as determined by the review. Solid lines display the median, the shaded area marks the interval between 10th and 90th percentile. Using moving average leads to systematically higher V̇O~2max~ values compared to binned time averages. Changing the averaging interval (in seconds or breaths) can lead to median changes in V̇O~2max~ as large as 5%,.](../plots/comparison.png)

Binned time averages systematically generate V̇O~2max~ values compared to moving averages. Decreasing the calculation interval to 5 seconds leads to a 3-4% median increase of V̇O~2max~ values. Notably on an individual level these increases may be lower (≈2%) or much higher (\>10%) than the median value.

Moving time and moving breath averages yield nearly identical values for V̇O~2max~ over all calculation intervals. This corresponds to the trained athletes reaching respiratory rates around 60 min^-1^ in the final minute of the exercise test (see Figure 4).

![Respiratory Rates peak around 60 min^-1^ in the ramp tests. The red segments correspond to the last minute before exhaustion of each individual (n = 72).](../plots/rr.png)

>>>>

Initial search yielded 7529 results of which 4364 remained after automated filtering and removal of duplicates (see flow diagram in @fig-flow). Out of the random sample (n = 500), 242 articles were included in the final analysis.

![Flow diagram for the systematic scoping review in accordance with the PRISMA 2020 Statement [@page2021]](../plots/flow_diagram.png){#fig-flow}

Reporting practices of the methodology of gas exchange measures differed widely across the literature (see @tbl-reporting). More than half (51.8%) of the articles did not report any information regarding their data processing strategy. One in twenty articles (5.4%) provided a rationale for their used strategy. Only a single article [@maturana2021] reported information regarding all the investigated criteria.

```{r}
#| echo: false
#| label: tbl-reporting
#| tbl-cap: "Percentage of studies that provided details on the different characteristics of oxygen uptake data processing. *only examined within the subgroup of studies using breath-by-breath measurements"
load("../data/reporting.Rda")
t <- purrr::map_dfr(reporting, scales::percent, accuracy = 0.1)

@@

knitr::kable(
  t,
  col.names = c("Metabolic cart", "Preprocessing", "Software", "Processing Strategy", "Reference"),
  align = "r"
)
```

Out of the authors that provided information and collected breath-by-breath measurements, most (79.5%) utilized binned averages to determine V̇O~2max~. Moving time averages, or breath-based averages were uncommon (see @fig-strategies). No study used methods of digital filtering to determine V̇O~2max~.

![Data strategies for processing breath-by-breath data in the reviewed literature (n = 88).](../plots/strategies.png){#fig-strategies}

@@

The calculation interval for time-based averages of mixing chamber and breath-by-breath devices ranged from 5 to 60 seconds (see @fig-duration). 30 second length intervals were most common to define V̇O~2max~, while authors also often employed shorter (10-20 s) and longer (60 s) periods.

![Total durations of the calculation interval of V̇O~2max~ in the reviewed studies.](../plots/duration_count.png){#fig-duration}
